---
title: "Individual differences in frontoparietal plasticity in humans"
author: "Boroshok et al. (2022)"
date: "06/20/2022"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

### RMarkdown has a known and common issue where Knitted files occassionally open in irrelevant applications (e.g., Skype, Calibre). If this happens, we recommend resetting your working directory and rendering this markdown file manually using the code below.
#setwd("~/Downloads/Materials for manuscript Individual differences in frontoparietal plasticity in humans/")
#render("plasticity.Rmd", output_format = "word_document") 

## Clean workspace
rm(list=ls())

# Uncomment these two lines if the script produces any errors.
#require('checkpoint')
#checkpoint("2021-04-27")

#### DEPENDENCIES ####
library(psych)
library(papaja)
library(dplyr)
# Install the stable development versions from GitHub
#devtools::install_github("crsh/papaja")

#### LOAD DATA ####

##################
### MRI sample ###
##################
plasticity_data <- read.csv("/Users/mackeylab-adm/Documents/BPD/data/plasticity_mri_sample.csv", header = T, sep = ",")

```

```{r mri sample results, include = FALSE}

# Relationships between baseline performance and learning gains
accuracy_mod <- lm(outofscanner_change_avg_pc ~ outofscanner_pre_avg_pc + age + sex, data = plasticity_data)
accuracy_mod <- apa_print(accuracy_mod, digits =3)
responsetime_mod <- lm(outofscanner_change_avg_rt ~ outofscanner_pre_avg_rt + age + sex, data = plasticity_data)
responsetime_mod <- apa_print(responsetime_mod, digits =3)

# Relationships between change in learning indices
learning_gains_mod <- lm(outofscanner_change_avg_pc ~ outofscanner_change_avg_rt + outofscanner_pre_avg_pc + outofscanner_pre_avg_rt + age + sex, data = plasticity_data)
learning_gains_mod <- apa_print(learning_gains_mod, digits =3)

# Relationships between learning gains and age, sex
age_accuracy_change_mod <- lm(outofscanner_change_avg_pc ~ age + outofscanner_pre_avg_pc, data = plasticity_data)
age_accuracy_change_mod <- apa_print(age_accuracy_change_mod, digits =3)
sex_accuracy_change_mod <- lm(outofscanner_change_avg_pc ~ sex + outofscanner_pre_avg_pc, data = plasticity_data)
sex_accuracy_change_mod <- apa_print(sex_accuracy_change_mod, digits =3)

age_responsetime_change_mod <- lm(outofscanner_change_avg_rt ~ age + outofscanner_pre_avg_rt, data = plasticity_data)
age_responsetime_change_mod <- apa_print(age_responsetime_change_mod, digits =3)
sex_responsetime_change_mod <- lm(outofscanner_change_avg_rt ~ sex + outofscanner_pre_avg_rt, data = plasticity_data)
sex_responsetime_change_mod <- apa_print(sex_responsetime_change_mod, digits =3)

# Normality of training measure
training_peak_shapiro_mod <- shapiro.test(plasticity_data$training_highestnback)
training_peak_shapiro_mod <- apa_print(training_peak_shapiro_mod, digits =3)


### Main models: VTA connectivity and accuracy
VTA_left_LPFC_accuracy_mod <- lm(outofscanner_change_avg_pc ~ L_lPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_pc, data = plasticity_data)
VTA_left_LPFC_accuracy_mod <- apa_print(VTA_left_LPFC_accuracy_mod, digits =3)

VTA_right_LPFC_accuracy_mod <- lm(outofscanner_change_avg_pc ~ R_lPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_pc, data = plasticity_data)
VTA_right_LPFC_accuracy_mod <- apa_print(VTA_right_LPFC_accuracy_mod, digits =3)

VTA_MPFC_accuracy_mod <- lm(outofscanner_change_avg_pc ~ mPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_pc, data = plasticity_data)
VTA_MPFC_accuracy_mod <- apa_print(VTA_MPFC_accuracy_mod, digits =3)

VTA_parietal_accuracy_mod <- lm(outofscanner_change_avg_pc ~ parietal_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_pc, data = plasticity_data)
VTA_parietal_accuracy_mod <- apa_print(VTA_parietal_accuracy_mod, digits =3)

VTA_striatum_accuracy_mod <- lm(outofscanner_change_avg_pc ~ striatum_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_pc, data = plasticity_data)
VTA_striatum_accuracy_mod <- apa_print(VTA_striatum_accuracy_mod, digits =3)


# no relationship between baseline VTA connectivity and RT change
VTA_left_LPFC_change_accuracy_mod <- lm(outofscanner_change_avg_rt ~ L_lPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_rt, data = plasticity_data)
VTA_left_LPFC_change_accuracy_mod <- apa_print(VTA_left_LPFC_change_accuracy_mod, digits = 3)

VTA_right_LPFC_change_accuracy_mod <- lm(outofscanner_change_avg_rt ~ R_lPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_rt, data = plasticity_data)
VTA_right_LPFC_change_accuracy_mod <- apa_print(VTA_right_LPFC_change_accuracy_mod, digits = 3)

VTA_mPFC_change_accuracy_mod <- lm(outofscanner_change_avg_rt ~ mPFC_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_rt, data = plasticity_data)
VTA_mPFC_change_accuracy_mod <- apa_print(VTA_mPFC_change_accuracy_mod, digits = 3)

VTA_parietal_change_accuracy_mod <- lm(outofscanner_change_avg_rt ~ parietal_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_rt, data = plasticity_data)
VTA_parietal_change_accuracy_mod <- apa_print(VTA_parietal_change_accuracy_mod, digits = 3)

VTA_striatum_change_accuracy_mod <- lm(outofscanner_change_avg_rt ~ striatum_VTA_A_rsFC + age + sex + rest_A_motion + rest_A_vols + outofscanner_pre_avg_rt, data = plasticity_data)
VTA_striatum_change_accuracy_mod <- apa_print(VTA_striatum_change_accuracy_mod, digits = 3)


### Main models: Myelin map values and response time
myelin_left_LPFC_responsetime_mod <- lm(outofscanner_change_avg_rt ~ L_lPFC_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
myelin_left_LPFC_responsetime_mod <- apa_print(myelin_left_LPFC_responsetime_mod, digits =3)

myelin_right_LPFC_responsetime_mod <- lm(outofscanner_change_avg_rt ~ R_lPFC_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
myelin_right_LPFC_responsetime_mod <- apa_print(myelin_right_LPFC_responsetime_mod, digits =3)

myelin_MPFC_responsetime_mod <- lm(outofscanner_change_avg_rt ~ mPFC_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
myelin_MPFC_responsetime_mod <- apa_print(myelin_MPFC_responsetime_mod, digits =3)

myelin_parietal_responsetime_mod <- lm(outofscanner_change_avg_rt ~ parietal_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
myelin_parietal_responsetime_mod <- apa_print(myelin_parietal_responsetime_mod, digits =3)

myelin_striatum_responsetime_mod <- lm(outofscanner_change_avg_rt ~ striatum_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
myelin_striatum_responsetime_mod <- apa_print(myelin_striatum_responsetime_mod, digits =3)


striatum_baseline_myelin_responsetime_mod <- lm(outofscanner_pre_avg_rt ~ striatum_A_myelin + age + sex, data = plasticity_data)
striatum_baseline_myelin_responsetime_mod <- apa_print(striatum_baseline_myelin_responsetime_mod, digits =3)

left_LPFC_baseline_myelin_responsetime_mod <- lm(outofscanner_pre_avg_rt ~ L_lPFC_A_myelin + age + sex, data = plasticity_data)
left_LPFC_baseline_myelin_responsetime_mod <- apa_print(left_LPFC_baseline_myelin_responsetime_mod, digits =3)

right_LPFC_baseline_myelin_responsetime_mod <- lm(outofscanner_pre_avg_rt ~ R_lPFC_A_myelin + age + sex, data = plasticity_data)
right_LPFC_baseline_myelin_responsetime_mod <- apa_print(right_LPFC_baseline_myelin_responsetime_mod, digits =3)

mPFC_baseline_myelin_responsetime_mod <- lm(outofscanner_pre_avg_rt ~ mPFC_A_myelin + age + sex, data = plasticity_data)
mPFC_baseline_myelin_responsetime_mod <- apa_print(mPFC_baseline_myelin_responsetime_mod, digits =3)

parietal_baseline_myelin_responsetime_mod <- lm(outofscanner_pre_avg_rt ~ parietal_A_myelin + age + sex, data = plasticity_data)
parietal_baseline_myelin_responsetime_mod <- apa_print(parietal_baseline_myelin_responsetime_mod, digits =3)


# no relationship between baseline T1w/T2w and PC change
left_LPFC_baseline_myelin_change_accuracy_mod <- lm(outofscanner_change_avg_pc ~ L_lPFC_A_myelin + age + sex + outofscanner_pre_avg_pc, data = plasticity_data)
left_LPFC_baseline_myelin_change_accuracy_mod <- apa_print(left_LPFC_baseline_myelin_change_accuracy_mod, digits = 3)

right_LPFC_baseline_myelin_change_accuracy_mod <- lm(outofscanner_change_avg_pc ~ R_lPFC_A_myelin + age + sex + outofscanner_pre_avg_pc, data = plasticity_data)
right_LPFC_baseline_myelin_change_accuracy_mod <- apa_print(right_LPFC_baseline_myelin_change_accuracy_mod, digits = 3)

mPFC_baseline_myelin_change_accuracy_mod <- lm(outofscanner_change_avg_pc ~ mPFC_A_myelin + age + sex + outofscanner_pre_avg_pc, data = plasticity_data)
mPFC_baseline_myelin_change_accuracy_mod <- apa_print(mPFC_baseline_myelin_change_accuracy_mod, digits = 3)

parietal_baseline_myelin_change_accuracy_mod <- lm(outofscanner_change_avg_pc ~ parietal_A_myelin + age + sex + outofscanner_pre_avg_pc, data = plasticity_data)
parietal_baseline_myelin_change_accuracy_mod <- apa_print(parietal_baseline_myelin_change_accuracy_mod, digits = 3)

striatum_baseline_myelin_change_accuracy_mod <- lm(outofscanner_change_avg_pc ~ striatum_A_myelin + age + sex + outofscanner_pre_avg_pc, data = plasticity_data)
striatum_baseline_myelin_change_accuracy_mod <- apa_print(striatum_baseline_myelin_change_accuracy_mod, digits = 3)


visual_baseline_myelin_change_reactiontime_mod <- lm(outofscanner_change_avg_rt ~ V1_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
visual_baseline_myelin_change_reactiontime_mod <- apa_print(visual_baseline_myelin_change_reactiontime_mod, digits = 3)

motor_baseline_myelin_change_reactiontime_mod <- lm(outofscanner_change_avg_rt ~ premotor_A_myelin + age + sex + outofscanner_pre_avg_rt, data = plasticity_data)
motor_baseline_myelin_change_reactiontime_mod <- apa_print(motor_baseline_myelin_change_reactiontime_mod, digits = 3)
  
```

# Results

Our analyses included data from 46 adults (30 female, ages 18–25) who completed MRI scanning before and after fifty minutes of n-back training (Fig. 1). Five frontoparietal regions of interest (ROIs) were identified based on the n-back fMRI task during the pre-training scan (2-back > 1-back contrast): left and right lateral prefrontal cortex, bilateral medial prefrontal cortex, bilateral parietal cortex, and striatum (Fig. 52). We extracted two neuroimaging measures from each ROI: VTA resting-state functional connectivity and T1w/T2w ratio (see Methods section for processing parameters). We used linear models to examine how resting state functional connectivity (rsFC) between VTA and each ROI, and T1w/T2w ratios in each ROI, predict learning gains following training (change in accuracy and response time [RT] on the out-of-scanner pre- and post-training n-back task).

The VTA connectivity models included the following covariates: age, sex, baseline n-back task performance, motion during the baseline resting-state fMRI scan, and the number of volumes acquired during the baseline resting-state fMRI scan. The T1w/T2w ratio models included the following covariates: age, sex, and baseline n-back task performance. All results underwent FDR-correction in R for 28 tests (7 ROIs [5 task-based, 2 control]), 2 learning measures [accuracy, response time], and 2 neural measures [VTA rsFC, T1w/T2w ratio]. All statistical analyses were performed using R (version 4.05) and RStudio (version 1.4.1106) software (R Foundation for Statistical Computing, Vienna, Austria).


Working memory performance improved with training.

We considered two behavioral measures of learning: accuracy change and response time (RT) change on the out-of-scanner pre- and post-training n-back task (Fig. 1). Fifty minutes of training led to small but significant increases in accuracy and decreases in response times (Table 1, Fig. 2b). However, there was considerable variability in training gains among individuals (Fig. 2a–c). Individuals with lower baseline accuracy improved more on accuracy following training (`r accuracy_mod$full$outofscanner_pre_avg_pc`). Individuals with slower baseline response times showed swifter response times following training (`r responsetime_mod$full$outofscanner_pre_avg_rt`). Gains in accuracy were not associated with improvements in response times (`r learning_gains_mod$full$outofscanner_change_avg_rt`). Improvements in learning were not significantly associated with age (accuracy: `r age_accuracy_change_mod$full$age`; RT: `r age_responsetime_change_mod$full$age`) or sex (accuracy: `r sex_accuracy_change_mod$full$sex`; RT: `r sex_responsetime_change_mod$full$sex`), controlling for baseline performance. The highest n-back condition reached during training was not used in brain analyses because the distribution was significantly non-parametric (Shapiro-Wilk:`r training_peak_shapiro_mod$statistic`), due to a small number of participants reaching very high conditions (Fig. 2c).


Plasticity as potential

Stronger VTA connectivity at baseline predicted greater improvements in accuracy

We used functional connectivity between the VTA and the task-based FPS regions of interest as a proxy measure for dopamine system connectivity (Fig. 3a). Consistent with the hypothesis that greater strength of dopamine system connectivity is associated with greater learning, we found that stronger resting-state functional connectivity between the VTA and the bilateral LPFC at baseline predicted greater improvements in accuracy (Fig. 3b, c: left LPFC: `r VTA_left_LPFC_accuracy_mod$full$L_lPFC_VTA_A_rsFC`, p(FDR) = .23; right LPFC: `r VTA_right_LPFC_accuracy_mod$full$R_lPFC_VTA_A_rsFC`, p(FDR) = .056), controlling for baseline accuracy, age, sex, motion, and total number of volumes. The relationship between VTA-LPFC connectivity and accuracy gains survived FDR correction for 28 tests (7 ROIs, 2 learning measures, and 2 neural measures) for the right LPFC but not the left LPFC. There were no significant associations between accuracy gains and VTA connectivity with the mPFC (`r VTA_MPFC_accuracy_mod$full$mPFC_VTA_A_rsFC`), parietal cortex (`r VTA_parietal_accuracy_mod$full$parietal_VTA_A_rsFC`), or striatum (`r VTA_striatum_accuracy_mod$full$striatum_VTA_A_rsFC`). Further, there were no significant associations between VTA-FPS connectivity and changes in response times in any ROI (left LPFC: `r VTA_left_LPFC_change_accuracy_mod$full$L_lPFC_VTA_A_rsFC`, pFDR = 0.213; right LPFC: `r VTA_right_LPFC_change_accuracy_mod$full$R_lPFC_VTA_A_rsFC`, pFDR = 0.298; mPFC: `r VTA_mPFC_change_accuracy_mod$full$mPFC_VTA_A_rsFC`, pFDR = 0.298; parietal: `r VTA_parietal_change_accuracy_mod$full$parieatl_VTA_A_rsFC`, pFDR = 0.248; striatum: `r VTA_striatum_change_accuracy_mod$full$striatum_VTA_A_rsFC`, pFDR = 0.450). There were no significant associations between VTA connectivity and accuracy or response times at baseline.


Lower T1w/T2w ratios at baseline predicted greater improvements in response times

We used the ratio of T1w/T2w intensities as a proxy measure for myelination (Fig. 4a). Individuals with lower baseline T1w/T2w ratios in all five FPS regions of interest showed greater improvements in response times (Fig. 4b–f: left LPFC: `r myelin_left_LPFC_responsetime_mod$full$L_lPFC_A_myelin`, p(FDR) = .02; right LPFC: `r myelin_right_LPFC_responsetime_mod$full$R_lPFC_A_myelin`, p(FDR) = .02; mPFC: `r myelin_MPFC_responsetime_mod$full$mPFC_A_myelin`, p(FDR) = .078 ; parietal cortex: `r myelin_parietal_responsetime_mod$full$parietal_A_myelin`, p(FDR) = .02; and striatum: `r myelin_striatum_responsetime_mod$full$striatum_A_myelin`, p(FDR) = .02), while controlling for baseline response times, age, and sex. T1w/T2w ratios were not associated with accuracy improvement in any of the task-based ROIs (left LPFC: `r left_LPFC_baseline_myelin_change_accuracy_mod$full$L_lPFC_A_myelin`, pFDR = 0.223; right LPFC: `r right_LPFC_baseline_myelin_change_accuracy_mod$full$R_lPFC_A_myelin`, pFDR = 0.223; mPFC: `r mPFC_baseline_myelin_change_accuracy_mod$full$mPFC_A_myelin`, pFDR = 0.348; parietal: `r parietal_baseline_myelin_change_accuracy_mod$full$parietal_A_myelin`, pFDR = 0.261; striatum: `r striatum_baseline_myelin_change_accuracy_mod$full$striatum_A_myelin`, pFDR = 0.213). Lower baseline T1w/T2w ratios in striatum were related to faster response times at baseline (`r striatum_baseline_myelin_responsetime_mod$full$striatum_A_myelin`), while controlling for age and sex. Associations between T1w/T2w ratios and baseline response times in the other FPS regions of interest were not significant (left LPFC: `r left_LPFC_baseline_myelin_responsetime_mod$full$L_lPFC_A_myelin`; right LPFC: `r right_LPFC_baseline_myelin_responsetime_mod$full$R_lPFC_A_myelin`; mPFC: `r mPFC_baseline_myelin_responsetime_mod$full$mPFC_A_myelin`; parietal cortex: `r parietal_baseline_myelin_responsetime_mod$full$parietal_A_myelin`). T1w/T2w ratios were not associated with accuracy at baseline.


Sensitivity analysis

To examine the specificity of the predictions in the frontoparietal system, we examined two control ROIs that we did not expect to predict learning: primary visual and motor cortex. Resting-state functional connectivity between the VTA and visual and motor ROIs did not predict changes in accuracy or response time following training. Baseline T1w/T2w ratios in visual and motor ROIs were not associated with changes in accuracy. However, lower baseline T1w/T2w ratios were associated with greater improvements in response time (visual: `r visual_baseline_myelin_change_reactiontime_mod$full$V1_A_myelin`, pFDR = 0.01; motor: `r motor_baseline_myelin_change_reactiontime_mod$full$premotor_A_myelin`, pFDR = 0.01).


Plasticity as a process: Brain activation changes with training were small and not strongly associated with learning

We did not observe training-related changes in VTA connectivity, or in T1w/T2w ratios (Supplementary Table 1). We also did not observe training-related changes in functional activation (Supplementary Fig. 1). Brain changes were not associated with learning. Relationships among brain change measures and learning for all regions of interest are shown in Supplementary Fig. 2.
